---
title: "Spanish Wine Quality Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Spanish Wine Quality Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fedesoriano/spanish-wine-quality-dataset\>*

### Reference:

*\<fedesoriano. (April 2022). Spanish Wine Quality Dataset. Retrieved [Date Retrieved] from https://www.kaggle.com/datasets/fedesoriano/spanish-wine-quality-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r Load dataset}
# Load the necessary package
library(readr)

# Load dataset and capture parsing issues
wine_data <- read.csv("wines_SPA.csv", colClasses = c(
  winery = "character",
  wine = "character",
  year = "character",
  rating = "numeric",
  num_reviews = "numeric",
  country = "character",
  region = "character",
  type = "factor",
  body = "factor",
  acidity = "factor",
  price = "numeric"
))

# Check for parsing issues
problems(wine_data)

# Display the structure of the dataset
str(wine_data)

# View the first few rows of the dataset
head(wine_data)

# View the dataset in a separate viewer window
View(wine_data)
```

## Measures of Frequency
```{r MOF}
# Frequency tables for categorical variables
winery_freq <- table(wine_data$winery)
type_freq <- table(wine_data$type)

# Display the frequency tables
print("Winery Frequency:")
print(winery_freq)

print("Type Frequency:")
print(type_freq)
```

## Measures of Central Tendency
```{r MOCT}
# Mean, median, and mode for numeric variables
mean_rating <- mean(wine_data$rating, na.rm = TRUE)
median_rating <- median(wine_data$rating, na.rm = TRUE)
mode_rating <- as.numeric(names(sort(table(wine_data$rating), decreasing = TRUE)[1]))

mean_num_reviews <- mean(wine_data$num_reviews, na.rm = TRUE)
median_num_reviews <- median(wine_data$num_reviews, na.rm = TRUE)
mode_num_reviews <- as.numeric(names(sort(table(wine_data$num_reviews), decreasing = TRUE)[1]))

mean_price <- mean(wine_data$price, na.rm = TRUE)
median_price <- median(wine_data$price, na.rm = TRUE)
mode_price <- as.numeric(names(sort(table(wine_data$price), decreasing = TRUE)[1]))

# Display the measures of central tendency
print(paste("Mean Rating:", mean_rating))
print(paste("Median Rating:", median_rating))
print(paste("Mode Rating:", mode_rating))

print(paste("Mean Number of Reviews:", mean_num_reviews))
print(paste("Median Number of Reviews:", median_num_reviews))
print(paste("Mode Number of Reviews:", mode_num_reviews))

print(paste("Mean Price:", mean_price))
print(paste("Median Price:", median_price))
print(paste("Mode Price:", mode_price))
```

## Measures of Distribution
```{r MOD}
# Standard deviation, variance, range, and IQR for numeric variables
sd_rating <- sd(wine_data$rating, na.rm = TRUE)
var_rating <- var(wine_data$rating, na.rm = TRUE)
range_rating <- range(wine_data$rating, na.rm = TRUE)
iqr_rating <- IQR(wine_data$rating, na.rm = TRUE)

sd_num_reviews <- sd(wine_data$num_reviews, na.rm = TRUE)
var_num_reviews <- var(wine_data$num_reviews, na.rm = TRUE)
range_num_reviews <- range(wine_data$num_reviews, na.rm = TRUE)
iqr_num_reviews <- IQR(wine_data$num_reviews, na.rm = TRUE)

sd_price <- sd(wine_data$price, na.rm = TRUE)
var_price <- var(wine_data$price, na.rm = TRUE)
range_price <- range(wine_data$price, na.rm = TRUE)
iqr_price <- IQR(wine_data$price, na.rm = TRUE)

# Display the measures of distribution
print(paste("Standard Deviation of Rating:", sd_rating))
print(paste("Variance of Rating:", var_rating))
print(paste("Range of Rating:", paste(range_rating, collapse = " - ")))
print(paste("IQR of Rating:", iqr_rating))

print(paste("Standard Deviation of Number of Reviews:", sd_num_reviews))
print(paste("Variance of Number of Reviews:", var_num_reviews))
print(paste("Range of Number of Reviews:", paste(range_num_reviews, collapse = " - ")))
print(paste("IQR of Number of Reviews:", iqr_num_reviews))

print(paste("Standard Deviation of Price:", sd_price))
print(paste("Variance of Price:", var_price))
print(paste("Range of Price:", paste(range_price, collapse = " - ")))
print(paste("IQR of Price:", iqr_price))
```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numeric variables
cor_matrix <- cor(wine_data[, c("rating", "num_reviews", "price")], use = "complete.obs")

# Display the correlation matrix
print("Correlation Matrix:")
print(cor_matrix)
```

## ANOVA
```{r ANOVA}
# Load necessary packages
library(dplyr)
library(car)  # For Levene's Test
library(ggplot2)  # For visualizing residuals


# Convert year to numeric, handling "N.V." and other non-numeric values
wine_data$year <- ifelse(wine_data$year == "N.V.", NA, as.integer(wine_data$year))

# ANOVA: Effect of wine type on price
anova_type <- aov(price ~ type, data = wine_data)
summary(anova_type)

# ANOVA: Effect of region on price
anova_region <- aov(price ~ region, data = wine_data)
summary(anova_region)

```

## Plots
```{r Plots}
# Visualization of results
# Boxplot of price by type
ggplot(wine_data, aes(x = type, y = price)) +
  geom_boxplot() +
  labs(title = "Boxplot of Price by Wine Type", x = "Wine Type", y = "Price")

# Boxplot of price by region
ggplot(wine_data, aes(x = region, y = price)) +
  geom_boxplot() +
  labs(title = "Boxplot of Price by Region", x = "Region", y = "Price")


# Load necessary packages
library(ggplot2)
library(reshape2)


# Convert year to numeric, handling "N.V." and other non-numeric values
wine_data$year <- ifelse(wine_data$year == "N.V.", NA, as.integer(wine_data$year))

# Univariate Plots

# Histogram for price
ggplot(wine_data, aes(x = price)) +
  geom_histogram(binwidth = 50, fill = "blue", color = "black") +
  labs(title = "Histogram of Wine Prices", x = "Price", y = "Frequency")

# Boxplot for price
ggplot(wine_data, aes(y = price)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Wine Prices", y = "Price")

# Bar plot for type
ggplot(wine_data, aes(x = type)) +
  geom_bar(fill = "orange") +
  labs(title = "Bar Plot of Wine Types", x = "Wine Type", y = "Count")

# Multivariate Plots

# Scatter plot for rating vs. price
ggplot(wine_data, aes(x = rating, y = price)) +
  geom_point(color = "blue", alpha = 0.5) +
  labs(title = "Scatter Plot of Rating vs. Price", x = "Rating", y = "Price")

# Boxplot of price by type
ggplot(wine_data, aes(x = type, y = price)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Price by Wine Type", x = "Wine Type", y = "Price")

# Boxplot of price by region
ggplot(wine_data, aes(x = region, y = price)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Boxplot of Price by Region", x = "Region", y = "Price") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Correlation matrix heatmap

# Compute correlation matrix
cor_matrix <- cor(wine_data[, c("rating", "num_reviews", "price")], use = "complete.obs")

# Melt the correlation matrix into long format
melted_cor_matrix <- melt(cor_matrix)

# Plot heatmap
ggplot(melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Correlation Matrix Heatmap", x = "", y = "") +
  theme_minimal()
```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Load necessary packages
library(VIM)
library(naniar)

# Summary of missing values
summary(wine_data)

# Count of missing values in each column
sapply(wine_data, function(x) sum(is.na(x)))

# Percentage of missing values in each column
sapply(wine_data, function(x) mean(is.na(x)) * 100)

# Total number of missing values
sum(is.na(wine_data))

# Rows with missing values
missing_rows <- wine_data[!complete.cases(wine_data), ]
head(missing_rows)

# Visualization of missing values

# Using VIM package
aggr_plot <- aggr(wine_data, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, labels = names(wine_data), cex.axis = 0.7, gap = 3, ylab = c("Missing data", "Pattern"))

# Using naniar package
gg_miss_var(wine_data, show_pct = TRUE) +
  labs(title = "Missing Values by Variable")
```

## Data Cleaning
```{r Clean Data}
# Load necessary packages
library(dplyr)
library(tidyr)

# Remove rows with missing values
wine_data_clean <- wine_data %>%
  drop_na()

# Display the structure of the cleaned dataset
str(wine_data_clean)

# View the first few rows of the cleaned dataset
head(wine_data_clean)

# Total number of missing values
sum(is.na(wine_data_clean))

# Load necessary libraries
library(dplyr)

# Drop the 'year' and 'country' columns from the dataset
wine_data_clean <- wine_data_clean %>% select(-year, -country)

# Verify that the columns have been removed
str(wine_data_clean)
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary packages
library(caTools)

# Ensure reproducibility
set.seed(123)

# Create a partition index
split <- sample.split(wine_data_clean$price, SplitRatio = 0.8)

# Split the data into training and testing sets
train_data <- subset(wine_data_clean, split == TRUE)
test_data <- subset(wine_data_clean, split == FALSE)

# Display the structure of the training and testing sets
dim(test_data)
dim(train_data)
```

## Bootstrapping
```{r Bootstrapping}
# Load necessary packages
library(boot)

# Ensure reproducibility
set.seed(123)

# Function to calculate the statistic of interest
bootstrap_stat <- function(data, indices) {
  sample_data <- data[indices, ]  # Resample with replacement
  return(mean(sample_data$price)) # Calculate the mean of the price variable
}

# Perform bootstrapping
bootstrap_results <- boot(data = wine_data_clean, statistic = bootstrap_stat, R = 1000)

# Summary of bootstrap results
print(bootstrap_results)

# Plot the bootstrap distribution
plot(bootstrap_results)
```

## Training Different Models
```{r Model training}
# Ensure 'type' is a factor
wine_data_clean$type <- as.factor(wine_data_clean$type)

# Load necessary packages
library(caret)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index <- createDataPartition(wine_data_clean$type, p = 0.8, list = FALSE, times = 1)
train_data <- wine_data_clean[train_index, ]
test_data <- wine_data_clean[-train_index, ]

# Remove factor variables with only one level
single_level_factors <- sapply(train_data, function(x) is.factor(x) && length(unique(x)) == 1)
train_data <- train_data[, !single_level_factors]

# Ensure the same columns are removed from test_data
test_data <- test_data[, colnames(train_data)]

# Define the training control
train_control <- trainControl(method = "cv", number = 10)


# Decision Tree
dt_model <- train(type ~ ., data = train_data, method = "rpart", trControl = train_control)

# KNN Classification
knn_model <- train(type ~ ., data = train_data, method = "knn", trControl = train_control, tuneLength = 10)

# Display model results
print(dt_model)
print(knn_model)
```

## Performance Comparison
```{r Performance Comparison}
# Compare model performances using resampling
results <- resamples(list(
  Decision_Tree = dt_model,
  KNN = knn_model
))

# Summary of the results
summary(results)

# Boxplots of the results
bwplot(results)

# Dot plots of the results
dotplot(results)
```

## Saving Model
```{r Saving Model}
# Save the decision tree model to a file
saveRDS(dt_model, file = "./models/dt_model.rds")

# Load the saved decision tree model
loaded_dt_model <- readRDS("./models/dt_model.rds")

# Prepare new data for prediction (example data, adjust according to your dataset)
new_data <- data.frame(
  winery = "Teso La Monja",
  wine = "Tinto",
  rating = 4.8,
  num_reviews = 100,
  region = "Toro",
  body = factor(5, levels = levels(train_data$body)),
  acidity = factor(3, levels = levels(train_data$acidity)),
  price = 50.0
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_dt_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

